Black-hole ringdown as a probe of higher-curvature gravity theories

### 2022-03-31 papers/paper_alt_theor_bounds/paper_alt_theor_bounds.tex
hs: edited dcs section; move to eft section

Second, the scalar field does influence the response of BHs to linearized perturbations and thus
affects the BH's QNM spectra.
%
% The calculation of the QNMs in this theory is more complicated than in GR already for
% nonrotating BHs. The reason is due to
The coupling between scalar field and the Gauss-Bonnet invariant, which
manifests into a coupling between scalar perturbation and gravitational
perturbations of polar parity, which, e.g., breaks the familiar equivalence
between the QNM spectra of polar and axial gravitational perturbations of
Schwarzschild BHs in GR.
%
The QNMs on nonrotating BHs in EdGB gravity were first computed in
Refs.~\cite{Pani:2009wy,Blazquez-Salcedo:2016enn} and these calculations were
extended to leading-order in BH spin in Ref.~\cite{Pierini:2021jxd}.

@@

This theory is described by the action~\cite{Jackiw:2003pm},
%
\begin{equation} \label{eq:action_dcs}
    S_{\rm CS} = \frac{1}{16 \pi}
    \int \dV
    \left[
    R - \tfrac{1}{2}(\nabla \vartheta)^2
    + \tfrac{1}{4} \ell^{2}_{\rm CS} \, \vartheta \, {}^{*}RR
    \right],
\end{equation}
%
where $\vartheta$ is a pseudo-scalar field, which couples
to the Pontryagin density
${}^{*}RR = {}^{*}R^{\mu}{}_{\nu}{}^{\rho\sigma} R^{\nu}_{\mu\rho\sigma}$,
%
% \begin{equation}
%     {}^{*}RR = {}^{*}R^{\mu}{}_{\nu}{}^{\rho\sigma} R^{\nu}_{\mu\rho\sigma},
% \end{equation}
%
where ${}^{*}R^{\mu}{}_{\nu}{}^{\rho\sigma}$ is the dual of the Riemann tensor,
defined as
%
${}^{*}R^{\mu}{}_{\nu}{}^{\rho\sigma} =
\epsilon^{\mu}{}_{\nu\gamma\delta}
R^{\gamma\delta\rho\sigma} / 2$,
%
and where $\epsilon^{\mu}{}_{\nu\gamma\delta}$ is the Levi-Civita tensor.
%
The coupling strength between scalar field and the Pontryagin density is set by the
coupling constant $\ell_{\rm CS}$, with dimensions of length.

The theory admits the Schwarzschild BH solution of GR, but predicts
that spinning BHs have (secondary) scalar hair, which, unlike in
scalar-Gauss-Bonnet gravity, is dipolar~\cite{Yunes:2009hc,Konno:2009kg}.
%
Consequently, the leading-scalar radiation channel is quadrupolar making this
theory presently unconstrained by from the GWs from the inspiral of BH
binaries alone~\cite{Nair:2019iur,Perkins:2021mhb}.
%
However, the theory has been constrained by Ref~\cite{Silva:2020acr} through a
combination information of x-ray observations of the isolated neutron star PSR
J0030+0451~\cite{Lommen:2000yt,NANOGrav:2017wvv} by the
NICER~\cite{Riley:2019yda,Miller:2019cac} and gravitational-wave observation of
the binary neutron star event GW170817~\cite{TheLIGOScientific:2017qsa}.

Similarly to the case of scalar-Gauss-Bonnet gravity, the perturbations of nonrotating
BHs in dynamical Chern-Simons gravity also have a coupling between scalar and
metric perturbations. However, the coupling is between the scalar and gravitational perturbations
of axial parity~\cite{Yunes:2007ss,Cardoso:2009pk,Molina:2010fb,Wagle:2021tam}.

@@

This theory is given by the following action,
%
\begin{equation} \label{eq:action_eft}
    S_{\rm EFT} = \frac{1}{16 \pi}
    \int \dd^4x \sqrt{-g}
    \left[ R
    +
    {\sum}_{n \geqslant 2} \ell^{2n - 2} L^{(2n)}
    \right],
\end{equation}
%
where $\ell$ is some length scale, assumed to be small compares to the length
scale associated with a BH, i.e., $\ell / M \ll 1$, and
$L^{(2n)}$ are corrections to the Einstein-Hilbert term action that
introduce involving higher-order curvature tensors (with $2n$ metric
derivatives).

@@

The polarization of the GW signal (in the observer's frame) are,
%
\begin{equation}
h_+(\iota,\varphi_0;t ) - i h_\times(\iota,\varphi_0;t) = \sum_{l, m} {}_{-\!2}Y_{l m}(\iota,\varphi_0)\, h_{l m}(t)\,,
\end{equation}
%
where ${}_{-\!2}Y_{l m}(\iota,\varphi_0)$ are the $-2$ spin-weighted spherical harmonics described by standard angular dependence indices $(l,m)$.\footnote{Note that we use $l$ to denote the angular dependence index and $\ell$ to denote the coupling constant in Sec.~\ref{sec:review_parspec}.} As our baseline model, we use the computationally efficient (time-domain) multipolar waveform model for quasicircular spin-aligned BBH mergers described in \cite{Mihaylov:2021bpf} (henceforth referred to as \SEOB{}\footnote{This waveform model is available in \texttt{LALSuite} \cite{lalsuite} as the \texttt{SEOBNRv4HM\_PA} waveform approximant.}) which contains the modes, $(l, |m|)=(2,2),(2,1)$, $(3,3)$, $(4,4)$, and $(5,5)$~\cite{Cotesta:2018fcv,Mihaylov:2021bpf}. The model uses an effective-one-body approach combined with a post-adiabatic solution of the equations of motion to describe the inspiral-plunge waveform, $h_{l m}^\mathrm{insp-plunge}$. An accurate description of the merger is incorporated through calibration with NR simulations, as described in~\cite{Cotesta:2018fcv}, along with information for the merger and ringdown phases, from BH perturbation theory. The merger-ringdown waveform, $h_{l m}^\mathrm{merger-RD}$, is then stitched to inspiral-plunge waveform, $h_{l m}^\mathrm{insp-plunge}$ at a certain time $t = t^{\textrm{match}}_{l m}$, as
%
\begin{equation}
h_{l m}(t) = h_{l m}^\mathrm{insp-plunge}\, \theta(t_\mathrm{match}^{l m} - t) + h_{l m}^\mathrm{merger-RD}\,\theta(t-t_\mathrm{match}^{l m})\,,
\end{equation}
%
where $\theta(t)$ is the Heaviside step function. The merger-ringdown waveform is expressed as an exponentially damped sinusoid~\citep{Bohe:2016gbl,Cotesta:2018fcv,Mihaylov:2021bpf}
%
\begin{equation}
\label{RD}
h_{l m}^{\textrm{merger-RD}}(t) = \nu \ \tilde{A}_{l m}(t)\ e^{i \tilde{\phi}_{l m}(t)} \ e^{-i \sigma_{l m 0}(t-t^{\textrm{match}}_{l m})},
\end{equation}
%
where
%
\begin{align}
\sigma_{l m 0} &= {\rm Re}(\sigma_{l m 0}) + i \, {\rm Im}(\sigma_{l m 0})\,,
\nonumber \\
&= \omega_{l m 0} - i / \tau_{l m 0}\,,
\end{align}
%
are the complex frequencies of the fundamental ($n=0$ overtone) QNMs of the remnant BH.
%
The frequencies $\omega_{l m 0}$ and damping times $\tau_{l m 0}$ of the $(l,
m)$ QNM can be read off from the real and imaginary parts of $\sigma_{l m 0}$
respectively.
%
The functions $\tilde{A}_{l m}(t)$ and $\tilde{\phi}_{l m}(t)$ are defined in~\cite{Bohe:2016gbl,Cotesta:2018fcv}.

>>>>

Second, the scalar field does influence the response of BHs to linearized perturbations and thus
affects the BH's QNM spectra.
%
% The calculation of the QNMs in this theory is more complicated than in GR already for
% nonrotating BHs. The reason is due to
The coupling between scalar field and the Gauss-Bonnet invariant, which
manifests into a coupling between scalar perturbation and gravitational
perturbations of polar parity, which, e.g., breaks the familiar equivalence
between the QNM spectra of polar and axial gravitational perturbations (isospectrality) of
Schwarzschild BHs in GR.
%
The QNMs on nonrotating BHs in EdGB gravity were first computed in
Refs.~\cite{Pani:2009wy,Blazquez-Salcedo:2016enn} and these calculations were
extended to leading-order in BH spin in Ref.~\cite{Pierini:2021jxd}.

@@

This theory is described by the action~\cite{Jackiw:2003pm},
%
\begin{equation} \label{eq:action_dcs}
    S_{\rm CS} = \frac{1}{16 \pi}
    \int \dV
    \left[
    R - \tfrac{1}{2}(\nabla \vartheta)^2
    + \tfrac{1}{4} \ell^{2}_{\rm CS} \, \vartheta \, {}^{*}RR
    \right],
\end{equation}
%
where $\vartheta$ is a pseudoscalar field, which couples to the Pontryagin
density
${}^{*}RR = {}^{*}R^{\mu}{}_{\nu}{}^{\rho\sigma} R^{\nu}_{\mu\rho\sigma}$,
%
% \begin{equation}
%     {}^{*}RR = {}^{*}R^{\mu}{}_{\nu}{}^{\rho\sigma} R^{\nu}_{\mu\rho\sigma},
% \end{equation}
%
where ${}^{*}R^{\mu}{}_{\nu}{}^{\rho\sigma}$ is the dual of the Riemann tensor,
defined as
%
${}^{*}R^{\mu}{}_{\nu}{}^{\rho\sigma} =
\epsilon^{\mu}{}_{\nu\gamma\delta}
R^{\gamma\delta\rho\sigma} / 2$,
%
and where $\epsilon^{\mu}{}_{\nu\gamma\delta}$ is the Levi-Civita tensor.
%
As in scalar-Gauss-Bonnet theories, $\int \dV {}^{*}RR$ alone is a boundary term~\cite{Jackiw:2003pm},
hence cannot contribute to the field equations in four-dimensions.
%
However, this terms can modify the theory's field equation by coupling it to
$\vartheta$.
%
The coupling strength set by the coupling constant $\ell_{\rm CS}$, with
dimensions of length.

The theory admits same the garden-variety Schwarzschild BH solution of
GR~\cite{Jackiw:2003pm}, but admits non-Kerr solutions for spinning BHs~\cite{Yunes:2009hc,Konno:2009kg}.
%
These BHs carry secondary scalar hair, which, unlike in EdGB gravity, is
dipolar.
%
As a consequence, the leading-scalar radiative channel of BBH to lose energy is quadrupolar,
which affects the GW phase at 2PN order.
%
Deviations from GR at this PN order are poorly constrained with present GW
observations~\cite{LIGOScientific:2021sio}, making dCS gravity unconstrained by
analyses that rely on the inspiral portion of GWs from BBH
binaries~\cite{Nair:2019iur,Perkins:2021mhb}.
%
However, the theory has been constrained by Ref.~\cite{Silva:2020acr}, that found
%
$\ell_{\rm CS} \leqslant 8.5$~km,
%
by folding data from the x-ray observations of the isolated neutron star (NS)
PSR~J0030+0451~\cite{Lommen:2000yt,NANOGrav:2017wvv} by
NICER~\cite{Riley:2019yda,Miller:2019cac} and from the GW observation of the binary
neutron star (BNS) GW170817~\cite{TheLIGOScientific:2017qsa}.

The QNMs of the Schwarzschild BH in dCS gravity where studied in Refs.~\cite{Yunes:2007ss,Cardoso:2009pk,Molina:2010fb},
that found that scalar perturbations and gravitational perturbations of axial parity couple. (Contrast with EdGB gravity).
%
This yields in a breakdown of isospectrality. The QNM spectra of slowly-rotating BHs in dCS gravity was
studied in Ref.~\cite{Wagle:2021tam}.

@@

Our last example of modified gravity theories are the so-called effective field theories (EFT) of GR.
They are described by the action
%
\begin{equation} \label{eq:action_eft}
    S_{\rm EFT} = \frac{1}{16 \pi}
    \int \dd^4x \sqrt{-g}
    \left[ R
    +
    \sum_{n \geqslant 2} \ell^{2n - 2} L^{(2n)}
    \right]\,,
\end{equation}
%
where $\ell$ is a lengthscale, assumed to be small compares to the lengthscale
associated with a BH, i.e., $\ell / M \ll 1$, and $L^{(2n)}$ are corrections to
that introduce higher-order curvature tensors (with $2n$ metric derivatives).

@@

The polarization of the GW signal (in the observer's frame) are,
%
\begin{equation}
h_+(\iota,\varphi_0;t ) - i h_\times(\iota,\varphi_0;t) = \sum_{l, m} {}_{-\!2}Y_{l m}(\iota,\varphi_0)\, h_{l m}(t)\,,
\end{equation}
%
where ${}_{-\!2}Y_{l m}(\iota,\varphi_0)$ are the $-2$ spin-weighted spherical harmonics described by standard angular dependence indices $(l,m)$.\footnote{Note that we use $l$ to denote the angular dependence index and $\ell$ to denote the coupling constant in Sec.~\ref{sec:review_parspec}.} As our baseline model, we use the computationally efficient (time-domain) multipolar waveform model for quasicircular spin-aligned BBH mergers described in \cite{Mihaylov:2021bpf} (henceforth referred to as \SEOB{}\footnote{This waveform model is available in \texttt{LALSuite} \cite{lalsuite} as the \texttt{SEOBNRv4HM\_PA} waveform approximant.}) which contains the modes, $(l, |m|)=(2,2),(2,1)$, $(3,3)$, $(4,4)$, and $(5,5)$~\cite{Cotesta:2018fcv,Mihaylov:2021bpf}.
%
The model uses an effective-one-body (EOB) approach combined with a post-adiabatic solution of the equations of motion to describe the inspiral-plunge waveform, $h_{l m}^\mathrm{insp-plunge}$. An accurate description of the merger is incorporated through calibration with NR simulations, as described in~\cite{Cotesta:2018fcv}, along with information for the merger and ringdown phases, from BH perturbation theory. The merger-ringdown waveform, $h_{l m}^\mathrm{merger-RD}$, is then stitched to inspiral-plunge waveform, $h_{l m}^\mathrm{insp-plunge}$ at a certain time $t = t^{\textrm{match}}_{l m}$, as
%
\begin{equation}
h_{l m}(t) = h_{l m}^\mathrm{insp-plunge}\, \theta(t_\mathrm{match}^{l m} - t) + h_{l m}^\mathrm{merger-RD}\,\theta(t-t_\mathrm{match}^{l m})\,,
\end{equation}
%
where $\theta(t)$ is the Heaviside step function. The merger-ringdown waveform is expressed as an exponentially damped sinusoid~\citep{Bohe:2016gbl,Cotesta:2018fcv,Mihaylov:2021bpf}
%
\begin{equation}
\label{RD}
h_{l m}^{\textrm{merger-RD}}(t) = \nu \ \tilde{A}_{l m}(t)\ e^{i \tilde{\phi}_{l m}(t)} \ e^{-i \sigma_{l m 0}(t-t^{\textrm{match}}_{l m})},
\end{equation}
%
where
%
\begin{align}
\sigma_{l m 0} &= {\rm Re}(\sigma_{l m 0}) + i \, {\rm Im}(\sigma_{l m 0})\,,
\nonumber \\
&= \omega_{l m 0} - i / \tau_{l m 0}\,,
\end{align}
%
are the complex frequencies of the fundamental ($n=0$ overtone) QNMs of the remnant BH.
%
The frequencies $\omega_{l m 0}$ and damping times $\tau_{l m 0}$ of the $(l,
m)$ QNM can be read off from the real and imaginary parts of $\sigma_{l m 0}$
respectively.
%
The functions $\tilde{A}_{l m}(t)$ and $\tilde{\phi}_{l m}(t)$ are defined in~\cite{Bohe:2016gbl,Cotesta:2018fcv}.
